Experience shapes individual foraging specialization and success in a virtual predator-prey system

### 2023-06-19 manuscript/Chapter2-manuscript.Rmd
minor edits to the MS .rmd suggested by Clint

There is currently a lack of consensus on the fitness advantages of specialized vs flexible foraging strategies [@Phillips.etal2017]. Some studies have found greater success for individuals specializing on a foraging tactic [@Pintor.etal2014;@Patrick.Weimerskirch2014a;@vandenBosch.etal2019] while others have found that individuals using flexible foraging tactics have greater success [@Paull.etal2012;@Santoro.etal2019;@Manlick.etal2021]. This discrepancy may in part be due to the limited information we have on the ecological contexts favoring individual specialization over flexibility. However, many studies also show that both type of foragers can attain equal success depending on timescales, and suggest that fluctuating resources throughout a predator's lifetime may be key in shaping individual foraging specialization [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015;@Phillips.etal2017;@Courbin.etal2018]. It is hypothesized that when resources are predictable (or stable), individual specialists should have higher capture rates by reducing the energy and time required to search for and handle prey. In contrast, individual generalists should benefit from adjusting to fluctuating resources by reducing the consequences of uncertainty [@dall2010]. While it is clear that predators can match their hunting behaviour and locomotion to their prey [@McGhee.etal2013a;@Bro-Jorgensen2013;@Chang.etal2017;@Szopa-Comley.Ioannou2022], the important question remaining is how. Given the dynamic nature of predator-prey interactions, it is unlikely that evolution alone could explain such behavioural adjustments by predators, because complex hunting strategies require practice to be effective [@Kelley.Magurran2011]. Uncovering the role of experience in shaping direct predator-prey interactions would help to predict the behavioural decisions and the success of predators when they are hunting.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and the prey. We recently showed, using an online predator-prey videogame called *Dead by Daylight* (*DBD*), that virtual systems can overcome this challenge [@FraserFranco.etal2022]. In this dynamic virtual system, four prey players need to forage for resources while avoiding predation by a fifth player. The prey can contribute to the group's success by healing or helping others escape from the predator, or be selfish and attempt to escape by themselves [@Cere.etal2021]. The predators' hunting tactics and success are driven by the movement of the prey [@FraserFranco.etal2022]. Predators should thus benefit from learning how the prey behave to be successful. Other virtual predator-prey studies show that predation regimes drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al. in prep). These examples outline how virtual systems could be useful in identifying general ecological patterns, which in turn might help to advance predator-prey research. They also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems, including *DBD*, offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we evaluated how hunting experience shapes predator foraging specialization and success using data from players in the online videogame *Dead by Daylight*. First, we investigate how predators develop their individual hunting expertise. We hypothesize that predators should differ in the development of their expertise if they encountered varying levels of difficulty with the prey that they pursued. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we test the hypothesis that experience will shape individual foraging specialization. We define foraging tactics as the mean movement speed of the predator along a slow-fast continuum, and the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (e.g. always using a fast tactic). If experience reduces the costs of switching between low and high speeds as predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then the efficiency of a specific tactic should increase with experience and everyone should instead specialize in their speed. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) would be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, both processes may operate at the same time if behavioural interactions with the prey vary among predators, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in their speed, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success probably because they learned from experience how to adjust their hunting style to their prey.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018], while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@Szopa-Comley.Ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (MFF personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@McGhee.etal2013a]. Indeed, 44% of the predator population altered its behaviour with experience to different degrees, and 12% even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.

@@

A general assumption of predator-prey studies is that predators maximize success by matching their tactic to their prey [@abramsEvolutionPredatorPreyInteractions2000], yet, it has remained largely unknown whether this results from predators learning how to hunt their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found support of our prediction that prey were driving individual differences in expertise in a predator population. We also confirmed that experience and predator-prey encounters drove the patterns of individual foraging specialization and flexibility. Our results suggest that predators learned with experience, as their success increased and their speed remained matched the speed of their prey. Even though individuals were not all equally successful, both specialized and flexible tactics achieved similar hunting success. One limitation of our study was that we couldn't monitor all the matches of the prey, which prevented us from assessing their responses to the predator. Hence, future studies should aim at monitoring reciprocal behavioural dynamics through time, which may reveal important insights on the mechanisms driving stable equilibria in predator-prey systems. Lastly, virtual systems are increasingly recognized among ecologists as being useful to test hypotheses on consumer-resource interactions [@Barbe.etal2020;@Beauchamp2020;@Cere.etal2021;@FraserFranco.etal2022]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

@@

We warmly thank Behaviour Interactive Inc. for our collaboration which helps in advancing predator-prey science. We thank Julien Céré who managed the communication between us and the company, and Marine Dupuy who helped us prepare the data used in this study. We thank Alastair J. Wilson for insightful comments on the results. We also thank all members of Pierre-Olivier Montiglio's laboratory who provided useful feedback on earlier versions on this work. This work was supported by an NSERC postgraduate doctoral scholarship (569716-2022).

>>>>

There is currently a lack of consensus on the fitness advantages of specialized vs flexible foraging strategies [@Phillips.etal2017]. Some studies have found greater success for individuals specializing on a foraging tactic [@Pintor.etal2014;@Patrick.Weimerskirch2014a;@vandenBosch.etal2019] while others have found that individuals using flexible foraging tactics have greater success [@Paull.etal2012;@Santoro.etal2019;@Manlick.etal2021]. This discrepancy may in part be due to the limited information we have on the ecological contexts favoring individual specialization over flexibility. However, many studies also show that both type of foragers can attain equal success depending on timescales, and suggest that fluctuating resources throughout a predator's lifetime may be key in shaping individual foraging specialization [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015;@Phillips.etal2017;@Courbin.etal2018]. It is hypothesized that when resources are predictable (or stable), individual specialists should have higher capture rates by reducing the energy and time required to search for and handle prey. In contrast, individual generalists should benefit from adjusting to fluctuating resources by reducing the consequences of uncertainty [@dall2010]. While it is clear that predators can match their hunting behaviour and locomotion to their prey [@McGhee.etal2013a;@Bro-Jorgensen2013;@Chang.etal2017;@Szopa-Comley.Ioannou2022], the important question remaining is how. Given the dynamic nature of predator-prey interactions, it is likely that ontogenic behavioural changes in predators play an important role, because complex hunting strategies require practice to be effective [@Kelley.Magurran2011]. Uncovering the role of experience in shaping direct predator-prey interactions would help to predict the behavioural decisions and the success of predators when they are hunting.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and the prey. We recently showed, using an online predator-prey videogame called *Dead by Daylight* (*DBD*), that virtual systems can overcome this challenge [@FraserFranco.etal2022]. In this dynamic virtual system, four prey players need to forage for resources while avoiding predation by a fifth player. The prey can contribute to the group's success by healing or helping others escape from the predator, or be selfish and attempt to escape by themselves [@Cere.etal2021]. The predators' hunting tactics and success are driven by the movement of the prey [@FraserFranco.etal2022]. Predators should thus benefit from learning how the prey behave to be successful. Other virtual predator-prey studies show that predation regimes drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al., public communications and unpublished manuscript). These examples outline how virtual systems could be useful in identifying general ecological patterns, which in turn might help to advance predator-prey research. They also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, virtual systems, including *DBD*, offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we evaluated how hunting experience shapes predator foraging specialization and success using data from players in the online videogame *Dead by Daylight*. First, we investigate how predators develop their individual hunting expertise. We hypothesize that predators should differ in the development of their expertise if they encountered varying levels of difficulty with the prey that they pursued. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we test the hypothesis that experience will shape individual foraging specialization. We define foraging tactics as the mean movement speed of the predator along a slow-fast continuum, and the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (e.g. always using a fast tactic). If experience reduces the costs of switching between low and high speeds as predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then the efficiency of a specific tactic should increase with experience and everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) would be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, both processes may operate at the same time if behavioural interactions with the prey vary among predators, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success probably because they learned from experience how to adjust their hunting style to their prey.

@@

The predator population maintained its speed and degree of specialization throughout the study period, and individuals differed only slightly in their average speeds. However, differences in specialization among individuals were important and increased slightly with experience. Predators hunting at higher speeds specialized in this tactic and tended to encounter groups of prey with similar speeds, while those hunting at slower speeds were flexible and tended to encounter varying groups of prey. On one hand, this pattern could arise if some predators adjusted their strategy to their prey, because fast-paced hunting is a specialized tactic suited for prey that use rapid evasive movements [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018], while a slower and flexible tactic can be useful to minimize the consequences of uncertainty when prey escape unpredictably [@Szopa-Comley.Ioannou2022]. However, we observed that 44% of the predator population showed close to no change in specialization with experience. Since these individuals were already specialized as novices (Fraser Franco, personal observation), there would be no point in them changing their tactic with experience if they were already successful, or there could have been costs to switch their strategy if they more often encountered faster prey. This could explain why the behavioural interactions between the predator and the prey remained stable across the predator's experience. Yet, because prey can also learn how to avoid predation [@kelley2003;@turner2006;@mccomb2011], we believe that the predator-prey phenotype matching more likely emerged from reciprocal adjustments in speed by predators and prey as they interacted [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@McGhee.etal2013a]. Indeed, 44% of the predator population altered its behaviour with experience to different degrees, and 12% even switched between flexibility and specialization. Hence, if the prey also learned through repeated interactions with the predators, it is possible that experience contributed in stabilizing the system as both were adusting to each other, similar to Red Queen dynamics [@brockhurst2014]. Our results appear to support this since many individuals were shifting their behavioural niche while the population niche remained stable across experience.

@@

A general assumption of predator-prey studies is that predators maximize success by matching their tactic to their prey [@abramsEvolutionPredatorPreyInteractions2000], yet, it has remained largely unknown whether this results from predators learning how to hunt their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where interactions were directly monitored, we found support of our prediction that prey were driving individual differences in expertise in a predator population. We also confirmed that experience and predator-prey encounters drove the patterns of individual foraging specialization and flexibility. Our results suggest that predators learned with experience, as their success increased and their speed remained matched to the speed of their prey. Even though individuals were not all equally successful, both specialized and flexible hunters achieved similar success. One limitation of our study was that we couldn't monitor all the matches of the prey, which prevented us from assessing their responses to the predator. Hence, future studies should aim at monitoring reciprocal behavioural dynamics through time, which may reveal important insights on the mechanisms driving stable equilibria in predator-prey systems. Lastly, virtual systems are increasingly recognized among ecologists as being useful to test hypotheses on consumer-resource interactions [@Barbe.etal2020;@Beauchamp2020;@Cere.etal2021;@FraserFranco.etal2022]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

@@

We warmly thank Behaviour Interactive Inc. for our collaboration which helps in advancing predator-prey science. We thank Julien Céré who managed the communication between us and Behaviour Interactive Inc., and Marine Dupuy who helped us prepare the data used in this study. We thank Alastair J. Wilson for insightful comments on the results. We also thank all members of Pierre-Olivier Montiglio's laboratory who provided useful feedback on earlier versions on this work. This work was supported by an NSERC postgraduate doctoral scholarship (569716-2022).
